---
title:  Final Project
output:
  html_document:
     css: html-md-01.css
     fig_caption: yes
     highlight: haddock
     number_sections: yes
     theme: flatly
     toc: yes
     toc_float: true
     collapsed: no
---

```{r set-options, echo=FALSE}
options(width = 105)
knitr::opts_chunk$set(dev='png', dpi=300, cache=TRUE)
pdf.options(useDingbats = TRUE)
```

# Introduction #

In this project, I analyze precipitation trends in Mexican tropical forests using the TerraClimate dataset. TerraClimate comprises monthly climate and climatic water balance for the globe from 1958 to 2015 in a spatial resolution of ~4 km. 
First, I downloaded the precipitation nc file from the Climatology Lab website. I read the dataset and visualize the first slice with the following R code: 

###Load packages 

```{r load_packages, message=FALSE}
 # load packages
library(ncdf4)
library(RNetCDF)
library(ncdf.tools)
library(lattice)
library(RColorBrewer)
library(latticeExtra)
library(sf)
library(raster)
library(rasterVis)
library(ggplot2)
library(ggthemes)
```

###Read data

Printing the nc file provides all the information about the dataset. 

```{r}
# set path and filename
ncpath <- "~/GEOG590/"
ncname <- "agg_terraclimate_ppt_1958_CurrentYear_GLOBE.nc"  
ncfname <- paste(ncpath, ncname, sep="")
# open a netCDF file
ncin <- nc_open(ncfname)
print(ncin)
```

###Get latitude, longitude and time
  
```{r}
# get longitude and latitude
lon <- ncvar_get(ncin,"lon")
nlon <- dim(lon)
head(lon)
lat <- ncvar_get(ncin,"lat")
nlat <- dim(lat)
head(lat)
print(c(nlon,nlat))

# get time
time <- ncvar_get(ncin,"time")
head(time); tail(time)
tunits <- ncatt_get(ncin,"time","units")
nt <- dim(time)
head(time); tail(time)
nt

# convert time -- split the time units string into fields
tustr <- strsplit(tunits$value, " ")
ptime <- convertDateNcdf2R(time, unlist(tustr)[1], origin = as.POSIXct(unlist(tustr)[3], 
                                                                       tz = "UTC"), time.format = "%Y-%m-%d")
head(time); tail(time)
head(ptime); tail(ptime)

# get data
dname <- "ppt" 
ppt_array <- ncvar_get(ncin,dname)
# get attributes of variable
dname <- "ppt" 
dlname <- ncatt_get(ncin,dname,"long_name")
dunits <- ncatt_get(ncin,dname,"units")
fillvalue <- ncatt_get(ncin,dname,"_FillValue")
print(c(dlname$value, dunits$value, fillvalue$value))
dim(ppt_array)

# replace netCDF fill values with NA's
ppt_array[ppt_array==fillvalue$value] <- NA
length(na.omit(as.vector(ppt_array[,,1])))
```
  

###Plot a slice of the array

I used the shapefile of world countries to add an outline.

```{r}
# set path and shape file name (countries outlines)
shp_path <- "~/GEOG590/shp_files/ne_110m_admin_0_countries/"
shp_name <- "ne_110m_admin_0_countries.shp"
shp_file <- paste(shp_path, shp_name, sep="")
# read the shapefile
world_shp <- read_sf(shp_file)
world_outline <- as(st_geometry(world_shp), Class="Spatial")

```

I selected the first slice, but it is possible to visualize any month of the 60 years

```{r}
# levelplot of a slice of the array
n <- 1
ppt_slice <- ppt_array[,, n]
grid <- expand.grid(lon=lon, lat=lat)
cutpts <- c(0,5,20,50,100,500,1000,2000,3000)
plt <- levelplot(ppt_slice ~ lon * lat, data=grid, at=cutpts, cuts=11, pretty=T, 
                 col.regions=((brewer.pal(9,"Blues"))))
plt + latticeExtra::layer(sp.lines(world_outline, col="black", lwd=0.5))
```

I will get the beginning and ending year of the time series

```{r}
# get beginning year and ending year, number of years, and set nm
beg_date <- strsplit(as.character(ptime[1]), " ")
beg_yr <- as.numeric(unlist(strsplit(as.character(beg_date), "-"))[1])
end_date <- strsplit(as.character(ptime[nt]), " ")
end_yr <- as.numeric(unlist(strsplit(as.character(end_date), "-"))[1])
nyrs <- end_yr - beg_yr + 1
nm <- 12
print(c(beg_yr, end_yr, nyrs, nm))
```

# Data analysis and visualization #

Research conducted in this region suggest that precipitation during the dry season does not present substantial variation, but the rainy season months have shown a decrease in precipitation. To get an idea of monthly trends, I will plot the average precipitation per month through time. 

```{r}
# generate a decimal year ("YrMn") time coordinate
YrMn <- seq(beg_yr, end_yr+1-(1/12), by=(1/12))
head(YrMn); tail(YrMn)
# year
year <- as.integer(YrMn)
head(year); tail(year)
# month
month <- rep(seq(1, 12, by=1), nyrs)
month_names <- c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov", "Dec")
month <- rep(month_names, nyrs)
head(month); tail(month)
month <- factor(month, levels=c("Dec","Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov"))
str(month)
```

###Precipitation Trend in two locations

I am interested in studying climatic changes in cloud forest ecosystems. Thus, I will compare the precipitation trends of two different locations of cloud forest.

```{r}
# get indices of the grid cell closest to Santa Cruz Tepetotutla, Oaxaca (1)
tlon1 <- -96.5593; tlat1 <- 17.7390
j <- which.min(abs(lon-tlon1))
k <- which.min(abs(lat-tlat1))
print(c(j, lon[j], k, lat[k]))
```

Plot the time series:
```{r}
# get time series for the closest gridpoint
ppt_ts <- ppt_array[j, k, ]

# plot time series of grid point
plot(YrMn, ppt_ts, type="l", xlab="Year", ylab=dname, main=dlname$value, col="blue")
```


```{r}
# get indices of the grid cell closest to Coatepec, Veracruz (2)
tlon2 <- -96.9954; tlat2 <- 19.4583
j2 <- which.min(abs(lon-tlon2))
k2 <- which.min(abs(lat-tlat2))
print(c(j2, lon[j2], k2, lat[k2]))
```

Plot the time series:
```{r}
# get time series for the closest gridpoint
ppt_ts2 <- ppt_array[j2, k2, ]

# plot time series of grid point
plot(YrMn, ppt_ts2, type="l", xlab="Year", ylab=dname, main=dlname$value, col="blue")
```

Now, I will plot monthly trends using a multi-panel plot
```{r}
# make dataframe
ppt_df <- data.frame(YrMn, year, month, ppt_ts)
str(ppt_df)
head(ppt_df); tail(ppt_df)
```

```{r}
ggplot(data = ppt_df, aes(x=year, y=ppt_ts)) +
  geom_smooth(method = "loess", size=0.5, color="blue") +
  geom_line() + 
  facet_grid(month ~ .)+ 
  theme(aspect.ratio = (0.04 * (nm/2)))  
```

```{r}
# make dataframe
ppt_df2 <- data.frame(YrMn, year, month, ppt_ts2)
str(ppt_df)
head(ppt_df); tail(ppt_df)
```

```{r}
ggplot(data = ppt_df2, aes(x=year, y=ppt_ts2)) +
  geom_smooth(method = "loess", size=0.5, color="blue") +
  geom_line() + 
  facet_grid(month ~ .)+ 
  theme(aspect.ratio = (0.04 * (nm/2)))   
```

Next steps: analyze spatial variation. 